---
title: "Bay(es) Window"
author: "Michael Betancourt"
date: "April 2024"
date-format: "MMMM YYYY"
toc: true
number-sections: true
highlight: pygments
bibliography: window_inference.bib
crossref:
  lst-title: "Stan Program"
filters:
   - include-code-files
format:
  html:
    html-math-method: katex
    theme:
      - lux
      - custom.scss
    standalone: true
    embed-resources: true
    code-overflow: wrap
    linkcolor: "#B97C7C"
  pdf:
    keep-tex: true
    fig-width: 4
    fig-height: 4
    code-overflow: wrap
    monofontoptions:
      - Scale=0.5
knitr:
  opts_chunk:
    comment: ''
  opts_knit:
    global.par: TRUE
format-links: false
---

Exercise 22.10 of @MacKay:2003 introduces the problem of inferring the
dimensions of a window in the dark, using only the positions of stars seen
through it.  In this short case study I present a comprehensive Bayesian
analysis of this problem.

# The Room

Consider a small room with square walls, only one of which features a window
that allows us to see to the outside.  During the day this view allows us to
fully distinguish the geometry of the window, including not only its width and
height but also its position on the wall (@fig-daytime).  On a moonless night,
however, the only light we see is from a scattering of stars.  The edges of the
window are enveloped by the darkness and become indistinguishable from the unlit
wall (@fig-nighttime).


::: {#fig-setup layout="[-4, 44, -4, 44, -4]"}
![](figures/daytime/daytime){#fig-daytime}

![](figures/nighttime/nighttime){#fig-nighttime}

External light passing through a window allow us to differentiate the window
from the surrounding wall.  (a) During the day the bright sunlight allows us to
fully characterize the window geometry but (b) during a moonless night we are
limited to starlight which only partially informs the window geometry.
:::

While the starlight does not allow us to completely determine the geometry of
the window it does provide some information.  Our goal is to infer the window
geometries that are consistent with the stars visible on a given night.

To that end let's set up a coordinate system that describes the geometry of the
room, including both the wall and the window (@fig-coordinate-system).  Let's
say that we do know that the height and width of the wall are both
$2 \, L = 6$ meters.  We can then denote the left, right, bottom, and top edges
of the window with the variables $l$, $r$, $b$, and $t$ respectively.

![A natural coordinate system for this problem is centered on the center of the
wall.  The left $l$, right $r$, bottom $b$, and top $t$ edges of the window can
then be described by distances from this center.](
figures/coordinate_system/coordinate_system){
width=50% #fig-coordinate-system}

# The Observational Model

The data generating process here is relatively straightforward.  Given a
collection of stars we observe the positions of the stars that fall within the
breadth of the window.  More formally we have a
[selection process](https://betanalpha.github.io/assets/chapters_html/modeling_selection.html)
with a latent distribution of star positions and a deterministic selection
function that rejects all stars that are obscured by the wall.

Firstly let's assume that the horizontal and vertical positions of each star
that we could possibly see are uniformly distributed across the total breadth of
the wall,
$$
p(x, y) = \text{uniform}(x \mid -L, +L) \, \text{uniform}(y \mid -L, +L).
$$

The wall then defines a deterministic selection function for a star being
observed,
$$
p(z = 1 \mid x, y, l, r, b, t) = I_{(l, r)}(x) \, I_{(b, t)}(y).
$$

Finally the observed star positions are modeled as
\begin{align*}
p(x, y &\mid z = 1, l, r, b, t)
\\
&=
\frac{ p(x, y) \, p(z = 1 \mid x, y) }
{ \int \mathrm{d} x \, \mathrm{d} y \, p(x, y) \, p(z = 1 \mid x, y) }
\\
&=
\frac{ \text{uniform}(x \mid -L, +L) \, \text{uniform}(y \mid -L, +L) \,
  I_{(l, r)}(x) \, I_{(b, t)}(y) }
{ \int \mathrm{d} x \, \mathrm{d} y \,
  \text{uniform}(x \mid -L, +L) \, \text{uniform}(y \mid -L, +L) \,
  I_{(l, r)}(x) \, I_{(b, t)}(y) }
\\
&=
\frac{ \text{uniform}(x \mid -L, +L) \, I_{(l, r)}(x) }
{ \int \mathrm{d} x \, \text{uniform}(x \mid -L, +L) \, I_{(l, r)}(x) } \,
\frac{ \text{uniform}(y \mid -L, +L) \, I_{(b, t)}(y) }
{ \int \mathrm{d} y \, \text{uniform}(y \mid -L, +L) \, I_{(b, t)}(y) }
\end{align*}

To simplify this further we'll need to take advantage of two facts.  First the
uniform probability density function is defined by an indicator function of its
own,
$$
\text{uniform}(z \mid a, b) = \frac{ I_{(a, b)}(z) }{ | b - a | }.
$$
Second the product of two interval indicator functions is given by the indicator
function of the intersection of the two intervals,
$$
I_{(a, b)}(z) \, I_{(c, d)}(z) = I_{(a, b) \cap (c, d)}(z).
$$
Putting these two properties together gives
\begin{align*}
\text{uniform}(z \mid a, b) \, I_{(c, d)}(z)
&=
\frac{ I_{(a, b)}(z) }{ | b - a | } \, I_{(c, d)}(z)
\\
&=
\frac{ I_{(a, b)}(z) \, I_{(c, d)}(z) }{ | b - a | }
\\
&=
\frac{ I_{(a, b) \cap (c, d)}(z) }{ | b - a | }.
\end{align*}

Because $(l, r) \subset (-L, +L)$ we have
$$
(-L, +L) \cap (l, r) = (l, r)
$$
and consequently
\begin{align*}
\text{uniform}(x \mid -L, +L) \, I_{(l, r)}(x)
&=
\frac{ I_{(-L, +L) \cap (l, r)}(x) }{ 2 L }
\\
&=
\frac{ I_{(l, r)}(x) }{ 2 L }.
\end{align*}
Similarly
$$
\text{uniform}(x \mid -L, +L) \, I_{(b, t)}(y) = \frac{ I_{(b, t)}(y) }{ 2 L }.
$$

Substituting these two results into the observational model for a single star
finally gives
\begin{align*}
p(x, y &\mid z = 1, l, r, b, t)
\\
&=
\frac{ \text{uniform}(x \mid -L, +L) \, I_{(l, r)}(x) }
{ \int \mathrm{d} x \, \text{uniform}(x \mid -L, +L) \, I_{(l, r)}(x) } \,
\frac{ \text{uniform}(y \mid -L, +L) \, I_{(b, t)}(y) }
{ \int \mathrm{d} y \, \text{uniform}(y \mid -L, +L) \, I_{(b, t)}(y) }
\\
&=
\frac{ I_{(l, r)}(x) / (2L) }{ \int \mathrm{d} x \, I_{(l, r)}(x) / (2L) } \,
\frac{ I_{(b, t)}(y) / (2L) }{ \int \mathrm{d} y \, I_{(b, t)}(y) / (2L) }
\\
&=
\frac{ I_{(l, r)}(x) }{ \int \mathrm{d} x \, I_{(l, r)}(x) } \,
\frac{ I_{(b, t)}(y) }{ \int \mathrm{d} y \, I_{(b, t)}(y) }
\\
&=
\frac{ I_{(l, r)}(x) }{ | r - l | } \,
\frac{ I_{(b, t)}(y) }{ | t - b | }.
\end{align*}

When we observe $N$ total stars with _independent_ positions $(x_{n}, y_{n})$
the total observational model becomes
\begin{align*}
p(x_{1}, y_{1}, \ldots, x_{N}, y_{N} \mid z = 1, l, r, b, t)
&=
\prod_{n = 1}^{N} \frac{ I_{(l, r)}(x_{n}) }{ | r - l | } \,
                  \frac{ I_{(b, t)}(y_{n}) }{ | t - b | }
\\
&=
\frac{ \prod_{n = 1}^{N} I_{(l, r)}(x_{n}) \, I_{(b, t)}(y_{n}) }
{ | r - l |^{N} \, | t - b |^{N} }.
\end{align*}

# The Likelihood Function

In theory the likelihood function for the window geometry is given by evaluating
the observational probability density function on a set observed positions,
\begin{align*}
\mathcal{L}(l, r, b, t)
&\propto
p(\tilde{x}_{1}, \tilde{y}_{1}, \ldots, \tilde{x}_{N}, \tilde{y}_{N} \mid
  z = 1, l, r, b, t)
\\
&\propto
\frac{
  \prod_{n = 1}^{N} I_{(l, r)}( \tilde{x}_{n}) \, I_{(b, t)}( \tilde{y}_{n})
}
{
  | r - l |^{N} \, | t - b |^{N}
}.
\end{align*}
This likelihood function, however, appears to be a bit ungainly to use in
practice because of the many discontinuities introduced by all of those
indicator functions.

Fortunately this issue is largely an illusion of how we have written the
likelihood function here; most of these indicator functions are in fact
redundant and can be removed entirely.  With a little bit of work we can
eliminate these redundancies and simplify the likelihood function into a form
that is much easier to wield in practice.

The key to this simplification is to manipulate the indicator functions over the
varying star positions into indicator functions over the common window position
parameters.  In general the indicator function $I_{(a, b)}(z)$ gives non-zero
outputs only when
$$
a < z < b.
$$
This sequence of inequalities, however, can also be written as three separate
inequalities,
\begin{align*}
a &< z
\\
z &< b
\\
a &< b.
\end{align*}
In turn these separate inequalities define three indicator functions for the
interval parameters $a$ and $b$,
\begin{align*}
I_{(-\infty, z)}(a)
\\
I_{(z, +\infty)}(b)
\\
I_{(-\infty, b)}(a).
\end{align*}
Consequently we can we can always write the indicator function $I_{(a, b)}(z)$
as a product of these three indicator functions,
$$
I_{(a, b)}(z)
=
I_{(-\infty, z)}(a) \, I_{(z, +\infty)}(b) \, I_{(-\infty, b)}(a).
$$
Again the first indicator function ensures that $a$ is less than $z$, the second
indicator function ensures that $b$ is larger than $z$, and the third indicator
function ensures that $a$ is less than $b$ which is required for $(a, b)$ to be
a well-defined interval.

This decomposition then allow us to write any product of indicator functions
evaluations as
\begin{align*}
\prod_{n = 1}^{N}
I_{(a, b)}(z_{n})
&=
\prod_{n = 1}^{N}
I_{(-\infty, z_{n})}(a) \, I_{(z_{n}, +\infty)}(b) \, I_{(-\infty, b)}(a)
\\
&=
\left[ \prod_{n = 1}^{N} I_{(-\infty, z_{n})}(a) \right] \,
\left[ \prod_{n = 1}^{N} I_{(z_{n}, +\infty)}(b) \right] \,
\left[ \prod_{n = 1}^{N} I_{(-\infty, b)}(a) \right]
\\
&=
\left[ \prod_{n = 1}^{N} I_{(-\infty, z_{n})}(a) \right] \,
\left[ \prod_{n = 1}^{N} I_{(z_{n}, +\infty)}(b) \right] \,
\left( I_{(-\infty, b)}(a) \right)^{N}
\\
&=
\left[ \prod_{n = 1}^{N} I_{(-\infty, z_{n})}(a) \right] \,
\left[ \prod_{n = 1}^{N} I_{(z_{n}, +\infty)}(b) \right] \,
I_{(-\infty, b)}(a).
\end{align*}

Now the overlap of the intervals $(-\infty, z_{n})$ is completely determined by
the smallest $z_{n}$,
$$
\bigcap_{n = 1}^{N} (-\infty, z_{n}) = (-\infty, \min(z_{n})).
$$
Consequently the first product reduces to
\begin{align*}
\prod_{n = 1}^{N} I_{(-\infty, z_{n})}(a)
&=
I_{ \cap_{n = 1}^{N} (-\infty, z_{n}) }(a)
\\
&=
I_{ (-\infty, \min(z_{n})) }(a).
\end{align*}
Similarly
$$
\bigcap_{n = 1}^{N} (z_{n}, +\infty) = (\max(z_{n}), +\infty).
$$
and
\begin{align*}
\prod_{n = 1}^{N} I_{(z_{n}, +\infty)}(b)
&=
I_{ \cap_{n = 1}^{N} (z_{n}, +\infty) }(b)
\\
&=
I_{ (\max(z_{n}), +\infty) }(b).
\end{align*}

This allows us to replace all of those indicator functions with just three
indicator functions,
\begin{align*}
\prod_{n = 1}^{N}
I_{(a, b)}(z_{n})
&=
\left[ \prod_{n = 1}^{N} I_{(-\infty, z_{n})}(a) \right] \,
\left[ \prod_{n = 1}^{N} I_{(z_{n}, +\infty)}(b) \right] \,
I_{(-\infty, b)}(a)
\\
&=
I_{ (-\infty, \min(z_{n})) }(a) \,
I_{ (\max(z_{n}), +\infty) }(b) \,
I_{(-\infty, b)}(a).
\end{align*}
For real-valued $z_{n}$ ties will occur with probability zero and we will have
$$
\min(z_{n}) < \max(z_{n})
$$
with probability one so that the last indicator function is also redundant.
Consequently we can write
$$
\prod_{n = 1}^{N}
I_{(a, b)}(z_{n})
=
I_{ (-\infty, \min(z_{n})) }(a) \,
I_{ (\max(z_{n}), +\infty) }(b).
$$

Applying this result twice our likelihood function takes on the much simpler
form
\begin{align*}
\mathcal{L}(l, r, b, t)
&\propto
\frac{
  \prod_{n = 1}^{N} I_{(l, r)}( \tilde{x}_{n}) \, I_{(b, t)}( \tilde{y}_{n})
}
{
  | r - l |^{N} \, | t - b |^{N}
}
\\
&\propto
\frac{
  \prod_{n = 1}^{N} I_{(l, r)}( \tilde{x}_{n}) \,
  \prod_{n = 1}^{N} I_{(b, t)}( \tilde{y}_{n})
}
{
  | r - l |^{N} \, | t - b |^{N}
}
\\
&\propto \;\,
\frac{
  I_{ (-\infty, \min(\tilde{x}_{n})) }(l) \,
  I_{ (\max(\tilde{x}_{n}), +\infty) }(r)
}
{
  | r - l |^{N}
}
\\
&\quad \cdot
\frac{
  I_{ (-\infty, \min(\tilde{y}_{n})) }(b) \,
  I_{ (\max(\tilde{y}_{n}), +\infty) }(t)
}
{
  | t - b |^{N}
}.
\end{align*}
The first indicator function keeps the left window boundary below all observed
$x$ positions while the second indicator function keeps the right window
boundary above all observed $y$ positions.  Similarly the last two indicator
functions ensure that the bottom window boundary is below, and the top window
boundary is above, all observed $y$ positions.  As the number of observations
$N$ grows the denominator pulls the likelihood function closer and closer to
the minimum and maximum $x$ and $y$ positions.

Because the likelihood function can always be written in terms of
$\min(\tilde{x}_{n})$, $\max(\tilde{x}_{n})$, $\min(\tilde{y}_{n})$, and
$\max(\tilde{y}_{n})$ regardless of what the individual positions are, these
extreme values are _sufficient statistics_ for this particular observational
model.

# Bayesian Inference

This simplified likelihood function is straightforward to implement in `Stan`,
allowing us to readily quantify posterior inferences for the window geometry
given observed star positions.

## Setup

As always we begin by setting up our local `R` environment.

```{r}
#| echo: !expr knitr::is_html_output()
#| eval: !expr knitr::is_html_output()
par(family="serif", las=1, bty="l",
    cex.axis=1, cex.lab=1, cex.main=1,
    xaxs="i", yaxs="i", mar = c(5, 5, 3, 1))

c_light <- c("#DCBCBC")
c_light_highlight <- c("#C79999")
c_mid <- c("#B97C7C")
c_mid_highlight <- c("#A25050")
c_dark <- c("#8F2727")
c_dark_highlight <- c("#7C0000")

c_light_teal <- c("#6B8E8E")
c_mid_teal <- c("#487575")
c_dark_teal <- c("#1D4F4F")

library(colormap)
```

```{r}
#| echo: !expr knitr::is_latex_output()
#| eval: !expr knitr::is_latex_output()
par(family="serif", las=1, bty="l",
    cex.axis=1, cex.lab=1, cex.main=1,
    xaxs="i", yaxs="i", mar = c(5, 5, 3, 5))

c_light <- c("#DCBCBC")
c_light_highlight <- c("#C79999")
c_mid <- c("#B97C7C")
c_mid_highlight <- c("#A25050")
c_dark <- c("#8F2727")
c_dark_highlight <- c("#7C0000")

c_light_teal <- c("#6B8E8E")
c_mid_teal <- c("#487575")
c_dark_teal <- c("#1D4F4F")

library(colormap)
disc_colors <- c("#FFFFFF", "#DCBCBC", "#C79999", "#B97C7C",
                 "#A25050", "#8F2727", "#7C0000")
```

In particular we'll need to load `RStan` and my recommended Hamiltonian Monte
Carlo analysis suite.

```{r}
#| warning: false
#| message: false
library(rstan)
rstan_options(auto_write = TRUE)            # Cache compiled Stan programs
options(mc.cores = parallel::detectCores()) # Parallelize chains
parallel:::setDefaultClusterOptions(setup_strategy = "sequential")

util <- new.env()
source('stan_utility_rstan.R', local=util)
```

## Explore Data

```{r}
#| eval: false
#| include: false

# Secret data simulation
left <- -1
right <- +1.5
bottom <- -0.5
top <- +2.0

stan_rdump(c('left', 'right', 'bottom', 'top'), "data/truth.data.R")

set.seed(48383992)

N <- 10
xs <- runif(N, left, right)
ys <- runif(N, bottom, top)

stan_rdump(c('N', 'xs', 'ys'), "data/stars.data.R")
```

Let's start by reading in the observed star positions.

```{r}
data <- read_rdump("data/stars.data.R")
```

The observed stars positions imply that the window is situated towards the top
right corner of the wall but the precise edges of the windows are hard to
resolve.

```{r}
L <- 3
par(mfrow=c(1, 1), mar=c(0, 0, 0, 0))

plot(0, type='n',
     xlim=c(-L, +L), ylim=c(-L, +L),
     axes=FALSE, ann=FALSE)
rect(-L, -L, +L, L, col = "black")
points(data$xs, data$ys, col="white", pch=8)
```

## Quantify Posterior Distribution

In addition to the requirement of a positive window size,
\begin{align*}
l &< r
\\
b &< t,
\end{align*}
the window geometry is constrained by the wall geometry,
\begin{align*}
-L &< l < +L
\\
-L &< r < +L
\\
-L &< b < +L
\\
-L &< t < +L.
\end{align*}

If we don't have any additional information about the position of the window
then our domain expertise is captured by the uniform prior density function
subject to these constraints,
\begin{align*}
p(l, r, b, t)
&=
p(l, r) \, p(b, t)
\\
&=\;\,
\text{uniform}(l \mid -L, +L) \, \text{uniform}(r \mid l, +L)
\\
&\quad \cdot
\text{uniform}(b \mid -L, +L) \, \text{uniform}(t \mid b, +L)
\end{align*}

With this prior model the posterior density function completely decouples into
independent posterior density functions for each of the two directions,
\begin{align*}
p(l, r, b, t &\mid
  \tilde{x}_{1}, \tilde{y}_{1}, \ldots, \tilde{x}_{N}, \tilde{y}_{N})
\\
&\propto
p(\tilde{x}_{1}, \tilde{y}_{1}, \ldots, \tilde{x}_{N}, \tilde{y}_{N}) \mid
  l, r, b, t) \,
p(l, r, b, t)
\\
&\propto \;
\frac{
  I_{ (-\infty, \min(\tilde{x}_{n})) }(l) \,
  I_{ (\max(\tilde{x}_{n}), +\infty) }(r)
}
{
  | r - l |^{N}
} \,
\\
&\quad \cdot
\frac{
  I_{ (-\infty, \min(\tilde{y}_{n})) }(b) \,
  I_{ (\max(\tilde{y}_{n}), +\infty) }(t)
}
{
  | t - b |^{N}
} \,
\\
&\quad \cdot
\text{uniform}(l \mid -L, +L) \, \text{uniform}(r \mid l, +L)
\\
&\quad \cdot
\text{uniform}(b \mid -L, +L) \, \text{uniform}(t \mid b, +L)
\\
&\propto \;
\frac{
  I_{ (-\infty, \min(\tilde{x}_{n})) }(l) \,
  I_{ (\max(\tilde{x}_{n}), +\infty) }(r) \,
}
{ | r - l |^{N} }
\\
&\quad \cdot
\text{uniform}(l \mid -L, +L) \, \text{uniform}(r \mid l, +L)
\\
&\quad \cdot
\frac{
  I_{ (-\infty, \min(\tilde{y}_{n})) }(b) \,
  I_{ (\max(\tilde{y}_{n}), +\infty) }(t)
}
{ | t - b |^{N} }
\\
&\quad \cdot
\text{uniform}(b \mid -L, +L) \, \text{uniform}(t \mid b, +L)
\\
&\propto \;\,
p(l, r \mid \tilde{x}_{1}, \tilde{y}_{1}, \ldots, \tilde{x}_{N}, \tilde{y}_{N})
\\
&\quad \cdot
p(b, t \mid \tilde{x}_{1}, \tilde{y}_{1}, \ldots, \tilde{x}_{N}, \tilde{y}_{N}).
\end{align*}


In particular we can immediately pushforward the four-dimensional posterior
density function into two-dimensional, marginal posterior density functions for
the horizontal and vertical geometries,
\begin{align*}
p(l, r \mid \tilde{x}_{1}, \tilde{y}_{1}, \ldots, \tilde{x}_{N}, \tilde{y}_{N})
&\propto \;
\frac{
  I_{ (-\infty, \min(\tilde{x}_{n})) }(l) \,
  I_{ (\max(\tilde{x}_{n}), +\infty) }(r) \,
}
{ | r - l |^{N} }
\\
&\quad \cdot
\text{uniform}(l \mid -L, +L) \, \text{uniform}(r \mid l, +L)
\end{align*}
and
\begin{align*}
p(b, t \mid \tilde{x}_{1}, \tilde{y}_{1}, \ldots, \tilde{x}_{N}, \tilde{y}_{N})
&\propto \;
\frac{
  I_{ (-\infty, \min(\tilde{y}_{n})) }(b) \,
  I_{ (\max(\tilde{y}_{n}), +\infty) }(t)
}
{ | t - b |^{N} }
\\
&\quad \cdot
\text{uniform}(b \mid -L, +L) \, \text{uniform}(t \mid b, +L).
\end{align*}
Plotting these marginal posterior density functions demonstrates how our
inferences push up against the constraints imposed by the minimum and maximum
star positions.

```{r}
par(mfrow=c(1, 2), mar=c(5, 5, 2, 1))

K <- 200
left_grid <- seq(-L, +L, 2 * L / (K - 1))
right_grid <- seq(-L, +L, 2 * L / (K - 1))

post_pds <- matrix(0, nrow=K, ncol=K)

for (k in 1:K) {
  for (kk in 1:K) {
    if (left_grid[k] > min(data$xs)) {
      post_pds[k, kk] <- 0
    } else if (right_grid[kk] < max(data$xs)) {
      post_pds[k, kk] <- 0
    } else {
      post_pds[k, kk] <- (right_grid[kk] - left_grid[k])**(-data$N)
    }
  }
}

disc_colors <- c("#FFFFFF", "#DCBCBC", "#C79999", "#B97C7C",
                 "#A25050", "#8F2727", "#7C0000")
cont_colors <- colormap(colormap=disc_colors, nshades=100)

image(left_grid, right_grid, post_pds, col=rev(cont_colors),
      xlim=c(-L, +L), xlab=c("Left"), ylim=c(-L, +L), ylab=c("Right"),
      main="Horizontal Dimensions")
abline(v=min(data$xs), lty=2, lwd=2, col="#DDDDDD")
text(0.75, -2.75, "min(x_n)", col="#DDDDDD")
abline(h=max(data$xs), lty=2, lwd=2, col="#DDDDDD")
text(2,  1.25, "max(x_n)", col="#DDDDDD")

K <- 200
bottom_grid <- seq(-L, +L, 2 * L / (K - 1))
top_grid <- seq(-L, +L, 2 * L / (K - 1))

post_pds <- matrix(0, nrow=K, ncol=K)

for (k in 1:K) {
  for (kk in 1:K) {
    if (bottom_grid[k] > min(data$ys)) {
      post_pds[k, kk] <- 0
    } else if (top_grid[kk] < max(data$ys)) {
      post_pds[k, kk] <- 0
    } else {
      post_pds[k, kk] <- (top_grid[kk] - bottom_grid[k])**(-data$N)
    }
  }
}

image(bottom_grid, top_grid, post_pds, col=rev(cont_colors),
      xlim=c(-L, +L), xlab=c("Bottom"), ylim=c(-L, +L), ylab=c("Top"),
      main="Vertical Dimensions")
abline(v=min(data$ys), lty=2, lwd=2, col="#DDDDDD")
text(1, -2.75, "min(y_n)", col="#DDDDDD")
abline(h=max(data$ys), lty=2, lwd=2, col="#DDDDDD")
text(2,  1.5, "max(y_n)", col="#DDDDDD")
```

If we want to extract additional information out of our posterior distribution
then we'll need to resort to posterior expectation values and the Markov chain
Monte Carlo estimators that approximate them.

The summary statistics $\min(\tilde{x}_{n})$, $\max(\tilde{x}_{n})$,
$\min(\tilde{y}_{n})$, and $\max(\tilde{y}_{n})$ make for natural retrodictive
summaries here.  Conveniently we can compute the posterior predictive summary
statistics directly in the `generated quantities` block.

```{.stan include="stan_programs/fit_window.stan" filename="fit\\\_window.stan"}
```

```{r}
#| warning: false
#| message: false
fit <- stan(file="stan_programs/fit_window.stan",
            data=data, seed=8438338,
            warmup=1000, iter=2024, refresh=0)
```

Conveniently our diagnostics show no serious indications of computational
infidelity.

```{r}
diagnostics <- util$extract_hmc_diagnostics(fit)
util$check_all_hmc_diagnostics(diagnostics)

samples <- util$extract_expectands(fit)
base_samples <- util$filter_expectands(samples,
                                       c('left', 'right',
                                         'bottom', 'top'))
util$check_all_expectand_diagnostics(base_samples)
```

None of the four summary statistics exhibit any appreciable retrodictive
tension, suggesting that our modeling assumptions are adequate.

```{r}
par(mfrow=c(2, 2), mar=c(5, 5, 2, 1))

util$plot_expectand_pushforward(samples[['min_x_pred']], 25, 'min_x',
                                baseline=min(data$xs))
util$plot_expectand_pushforward(samples[['max_x_pred']], 25, 'max_x',
                                baseline=max(data$xs))
util$plot_expectand_pushforward(samples[['min_y_pred']], 25, 'min_y',
                                baseline=min(data$ys))
util$plot_expectand_pushforward(samples[['max_y_pred']], 25, 'max_y',
                                baseline=max(data$ys))
```

The Markov chain Monte Carlo fit now allows us to examine marginal posterior
inferences for each individual parameter.  All four of the marginal posterior
distributions concentrate towards the hard edge imposed by the observed summary
statistics.

```{r}
par(mfrow=c(2, 2), mar=c(5, 5, 2, 1))

util$plot_expectand_pushforward(samples[['left']], 25,
                                'Left Window Edge', flim=c(-L, +L))
abline(v=min(data$xs), lty=2, lwd=2, col="#DDDDDD")
text(1, 1.4, "min(x_n)", col="#DDDDDD")

util$plot_expectand_pushforward(samples[['right']], 25,
                                'Right Window Edge', flim=c(-L, +L))
abline(v=max(data$xs), lty=2, lwd=2, col="#DDDDDD")
text(0.25, 1, "max(x_n)", col="#DDDDDD")

util$plot_expectand_pushforward(samples[['bottom']], 25,
                                'Bottom Window Edge', flim=c(-L, +L))
abline(v=min(data$ys), lty=2, lwd=2, col="#DDDDDD")
text(1, 1.4, "min(y_n)", col="#DDDDDD")

util$plot_expectand_pushforward(samples[['top']], 25,
                                'Top Window Edge', flim=c(-L, +L))
abline(v=max(data$ys), lty=2, lwd=2, col="#DDDDDD")
text(0.75, 1.5, "max(y_n)", col="#DDDDDD")
```

Perhaps the most useful benefit of our Markov chain Monte Carlo fit is that we
can visualize the inferred window geometry directly instead of having to
recreate it from its individual pieces.

First let's flatten the Markov chains into a single array for each parameter.

```{r}
left_samples <- c(samples[['left']], recursive=TRUE)
right_samples <- c(samples[['right']], recursive=TRUE)
bottom_samples <- c(samples[['bottom']], recursive=TRUE)
top_samples <- c(samples[['top']], recursive=TRUE)
```

Subsetting these flattened arrays selects out a selection of realized window
geometries consistent with the observed data.  Plotting these over against the
observed star positions puts our inferences into a direct geometric context.

```{r}
plot_idxs <- seq(1, 4096, 150)

disc_colors <- c("#B97C7C", "#A25050", "#8F2727", "#7C0000")
line_colors <- colormap(colormap=disc_colors,
                        nshades=length(plot_idxs))

par(mfrow=c(1, 1), mar=c(0, 0, 0, 0))

plot(0, type='n',
     xlim=c(-L, +L), ylim=c(-L, +L),
     axes=FALSE, ann=FALSE)
rect(-L, -L, +L, +L, col = "black")

points(data$xs, data$ys, col="white", pch=8)

for(s in seq_along(plot_idxs)) {
  plot_idx <- plot_idxs[s]
  lines(c(left_samples[plot_idx], right_samples[plot_idx],
          right_samples[plot_idx], left_samples[plot_idx],
          left_samples[plot_idx]),
        c(bottom_samples[plot_idx], bottom_samples[plot_idx],
          top_samples[plot_idx], top_samples[plot_idx],
          bottom_samples[plot_idx]),
        col=line_colors[s], lwd=2)
}
```

Finally let's compare our posterior inferences to the actual window geometry.

```{r}
truth <- read_rdump("data/truth.data.R")
```

```{r}
par(mfrow=c(1, 1), mar=c(0, 0, 0, 0))

plot(0, type='n',
     xlim=c(-L, +L), ylim=c(-L, +L),
     axes=FALSE, ann=FALSE)
rect(-L, -L, +L, +L, col = "black")

points(data$xs, data$ys, col="white", pch=8)

for(s in seq_along(plot_idxs)) {
  plot_idx <- plot_idxs[s]
  lines(c(left_samples[plot_idx], right_samples[plot_idx],
          right_samples[plot_idx], left_samples[plot_idx],
          left_samples[plot_idx]),
        c(bottom_samples[plot_idx], bottom_samples[plot_idx],
          top_samples[plot_idx], top_samples[plot_idx],
          bottom_samples[plot_idx]),
        col=line_colors[s], lwd=2)
}

lines(c(truth$left, truth$right, truth$right, truth$left, truth$left),
      c(truth$bottom, truth$bottom, truth$top, truth$top, truth$bottom),
      col="white", lwd=6)
lines(c(truth$left, truth$right, truth$right, truth$left, truth$left),
      c(truth$bottom, truth$bottom, truth$top, truth$top, truth$bottom),
      col=c_mid_teal, lwd=3)
```

Interesting in turns out that we have a bit of a void in the observed stars,
with the minimum $x$ position relatively far away from the true left window
edge.  Our posterior uncertainties, however, are able to account for this
fluctuation and ensure that we aren't lead astray.

# Conclusion

Inferring the dimensions of a window from the stars we see through it is an
excellent demonstration of narratively generative modeling in action.
Our understanding of the data generating process directly informs an
observational model and sets the context for constructing an initial prior
model.  Moreover the structure of that observational model immediately motivates
informative summary statistics on which we can build powerful posterior
retrodictive checks.

As a literal textbook exercise this problem is nice enough to admit a
substantial amount of analytic simplification.  That said many of most
interpretable insights are still obstructed by intractable calculations and
tools like Markov chain Monte Carlo are invaluable to the implementation of a
rich analysis.

# Acknowledgements {-}

A very special thanks to everyone supporting me on Patreon:
Adam Fleischhacker, Adriano Yoshino, Alessandro Varacca, Alexander Noll,
Alexander Petrov, Alexander Rosteck, Andrea Serafino, Andrew Mascioli,
Andrew Rouillard, Andrew Vigotsky, Ara Winter, Austin Rochford, Avraham Adler,
Ben Matthews, Ben Swallow, Benoit Essiambre, Bertrand Wilden, Bradley Kolb,
Brandon Liu, Brendan Galdo, Brynjolfur Gauti Jónsson, Cameron Smith,
Canaan Breiss, Cat Shark, Charles Naylor, Charles Shaw, Chase Dwelle,
Chris Jones, Christopher Mehrvarzi, Colin Carroll, Colin McAuliffe,
Damien Mannion, dan mackinlay, Dan W Joyce, Dan Waxman, Dan Weitzenfeld,
Daniel Edward Marthaler, Darshan Pandit, Darthmaluus , Dav Pe, David Galley,
David Wurtz, Denis Vlašiček, Doug Rivers, Dr. Jobo, Dr. Omri Har Shemesh,
Dylan Maher, Ed Cashin, Edgar Merkle, Eric LaMotte, Ero Carrera, Eugene O'Friel,
Felipe González, Fergus Chadwick, Finn Lindgren, Florian Wellmann,
Geoff Rollins, Guido Biele, Håkan Johansson, Hamed Bastan-Hagh, Hauke Burde,
Hector Munoz, Henri Wallen, hs, Hugo Botha, Ian, Ian Costley, idontgetoutmuch,
Ignacio Vera, Ilaria Prosdocimi, Isaac Vock, J, J Michael Burgess, jacob pine,
Jair Andrade, James C, James Hodgson, James Wade, Janek Berger, Jason Martin,
Jason Pekos, Jason Wong, jd, Jeff Burnett, Jeff Dotson, Jeff Helzner,
Jeffrey Erlich, Jesse Wolfhagen, Jessica Graves, Joe Wagner, John Flournoy,
Jonathan H. Morgan, Jonathon Vallejo, Joran Jongerling, JU, Justin Bois,
Kádár András, Karim Naguib, Karim Osman, Kejia Shi, Kristian Gårdhus Wichmann,
Lars Barquist, lizzie , LOU ODETTE, Luís F, Marcel Lüthi, Marek Kwiatkowski,
Mark Donoghoe, Markus P., Márton Vaitkus, Matt Moores, Matthew, Matthew Kay,
Matthieu LEROY, Mattia Arsendi, Maurits van der Meer, Michael Colaresi,
Michael DeWitt, Michael Dillon, Michael Lerner, Mick Cooney, N Sanders, N.S. ,
Name, Nathaniel Burbank, Nic Fishman, Nicholas Clark, Nicholas Cowie, Nick S,
Octavio Medina, Ole Rogeberg, Oliver Crook, Olivier Ma, Patrick  Kelley,
Patrick Boehnke, Pau Pereira Batlle, Peter Johnson, Pieter van den Berg , ptr,
Ramiro Barrantes Reynolds, Raúl Peralta Lozada, Ravin Kumar, Rémi , Rex Ha,
Riccardo Fusaroli, Richard Nerland, Robert Frost, Robert Goldman, Robert kohn,
Robin Taylor, Ryan Grossman, Ryan Kelly, S Hong, Sean Wilson, Sergiy Protsiv,
Seth Axen, shira, Simon Duane, Simon Lilburn, sssz, Stan_user, Stephen Lienhard,
Stew Watts, Stone Chen, Susan Holmes, Svilup, Tao Ye, Tate Tunstall,
Tatsuo Okubo, Teresa Ortiz, Theodore Dasher, Thomas Kealy, Thomas Siegert,
Thomas Vladeck, Tom McEwen, Tomáš Frýda, Tony Wuersch, Virginia Fisher,
Vladimir Markov, Wil Yegelwel, Will Farr, woejozney, yolhaj , yureq , Zach A,
Zad Rafi and Zhengchen Cai.

# References {-}

<div id="refs"></div>

# License {-}

A repository containing all of the files used to generate this chapter is
available on [GitHub](
https://github.com/betanalpha/quarto_chapters/tree/main/case_studies/window_inference).

The text and figures in this chapter are copyrighted by Michael Betancourt
and licensed under the CC BY-NC 4.0 license:

https://creativecommons.org/licenses/by-nc/4.0/
